---
title: "ST 518 Homework 8"
author: "Eric Warren"
date: "November 4, 2023"
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(scipen = 999)
```

```{r output setup, eval=FALSE, echo=FALSE}
# This code allows us to render a pdf document
rmarkdown::render("~/ST-518/Warren_ST 518 HW 8.Rmd", 
              output_format = "pdf_document", 
              output_options = list(
                toc = TRUE, 
                toc_depth = 3,
                number_sections = TRUE
                )
              )
```

# Problem 1

An experiment randomizes seeds from `t = 10` plants from an F2 generation of soybeans to `N = 30` homogeneous plots. The percentage protein content is measured in the seds from the plants produced in each plot with results below and in the file **“protein-content.dat”**.

First we are going to read in the data.
```{r read in problem 1 data}
library(tidyverse)
(protein <- read_table("protein-content.dat"))
```

Consider a random effects model for these protein contents:
$Y_{ij} = \mu + T_i + E_{ij}$.

## Part A

Give all distributional assumptions and limits on indices/subscripts i and j.

